---
title: "Calculate normal/mm ratios"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

### The code is fine, but the data cannot be confidently combined to calculate the ratio. 

Methods: 
1) Start with gene matrix
2) Convert into z-score
3) Calculate mean expression value for each gene (separately in HSC and MMRF)
4) Normal/MM ratio or MM/Normal ratio?
5) Barplot or boxplot

```{r libraries}
library(ggplot2)
library(RColorBrewer)
library(tidyr)
library(ggpubr)
```

Re-run the merging:
1) increase pool of genes used to calculate z-score
2) find a house-keeping gene similarly expressed between HSC and MM
3) use ComBat - sva package
```{r try (1) - ehh}
hsc <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/normals/gse17054_annot_log2.rds")
rownames(hsc)[rownames(hsc)=="TNFRSF17"] <- "BCMA"
rownames(hsc)[rownames(hsc)=="NCAM1"] <- "CD56"
rownames(hsc)[rownames(hsc)=="SDC1"] <- "CD138"
rownames(hsc)[rownames(hsc)=="TNFRSF13B"] <- "TACI"

ark <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/mm_pts/arkansas_annot_log2.rds")

rand.genes <- sample(rownames(hsc), 500)

genelist <- unique(c("BCMA", "CD38", "CD56", "CD138", "TACI", 
                 "SLAMF7", "CD19", "LY9", "ITGB7", "GPRC5D", 
                 "CD40", "CD44", "CD47", "BST2", "IGF1R", 
                 "ICAM1", "CD74", "CD81", "CD86", "CD200", "IGKC",
                 "FUT3", "CD70", "CD1D", rand.genes))
# subset genes: get indices
idxs <- sapply(genelist, function(x) grep(paste0("^", x, "$"), rownames(hsc)))
hsc.mat <- hsc[idxs, ]
idxs <- sapply(genelist, function(x) grep(paste0("^", x, "$"), rownames(ark)))
ark.mat <- ark[idxs, ]

#### now try z-score
hsc.zscore <- scale(hsc.mat, center=TRUE)
head(hsc.zscore)
# hist(hsc.zscore["BCMA", ], breaks=4)

ark.zscore <- scale(ark.mat, center=TRUE)
ark.zscore[1:5, 1:5]
# hist(ark.zscore["BCMA", ])

### calc means
hsc.means <- apply(hsc.zscore, 1, mean)
ark.means <- apply(ark.zscore, 1, mean) 

### calc mm-normal
ark.hsc <- NULL
for(g in names(hsc.means)) {
  mm.norm <- mmrf.means[g] - hsc.means[g]
  mmrf.hsc <- c(mmrf.hsc, mm.norm)
  
  mm.norm <- ark.means[g] - hsc.means[g]
  ark.hsc <- c(ark.hsc, mm.norm)
}

# hist(ark.hsc)
summary(ark.hsc)
# barplot(ark.hsc)

head(ark.hsc)

### keep 24 genes
genelist <- c("BCMA", "CD38", "CD56", "CD138", "TACI", 
                 "SLAMF7", "CD19", "LY9", "ITGB7", "GPRC5D", 
                 "CD40", "CD44", "CD47", "BST2", "IGF1R", 
                 "ICAM1", "CD74", "CD81", "CD86", "CD200", "IGKC",
                 "FUT3", "CD70", "CD1D")
# subset genes: get indices
idxs <- sapply(genelist, function(x) grep(paste0("^", x, "$"), names(ark.hsc)))
ark.hsc.genes24 <- ark.hsc[idxs]
# barplot(ark.hsc.genes24)

### plot
mat <- ark.hsc.genes24
df <- data.frame(Genes=names(mat), Diff=mat)
df$Color <- ifelse(df$Diff > 0, "#CA0020", "#0571B0") # red for positive difference, blue for negative difference
df$Genes <- reorder(df$Genes, df$Diff) # order by difference
## plot
ggplot(df, aes(x=Genes, y=Diff)) +
  geom_bar(stat="identity", fill=df$Color) +
  coord_flip() +
  theme_pubr() +
  xlab("") +
  ylab("Cancer - Normal (z-score)")
```

```{r try (2) - ehh, works?}
hsc <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/normals/gse17054_annot_log2.rds")
dim(hsc) # 20136 x 4
head(hsc)
rownames(hsc)[rownames(hsc)=="TNFRSF17"] <- "BCMA"
rownames(hsc)[rownames(hsc)=="NCAM1"] <- "CD56"
rownames(hsc)[rownames(hsc)=="SDC1"] <- "CD138"
rownames(hsc)[rownames(hsc)=="TNFRSF13B"] <- "TACI"
hsc.rowMeans <- rowMeans(hsc)
hsc.rowMeans[low.expr.genenames]

ark <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/mm_pts/arkansas_annot_log2.rds")
dim(ark) # 20316 x 1035
head(ark[, 1:5])
ark.rowMeans <- rowMeans(ark)
head(ark.rowMeans)
ark.rowMeans[low.expr.genenames]

diff.rowMeans <- sapply(names(ark.rowMeans), function(g) {
  return(ark.rowMeans[g] - hsc.rowMeans[g])
})

summary(diff.rowMeans)
sum(diff.rowMeans < 1, na.rm=TRUE)
low.expr.genes <- diff.rowMeans[which(diff.rowMeans < 1)] ## 26 genes, which are house-keeping?
low.expr.genenames <- sapply(strsplit(names(low.expr.genes), "\\."), '[', 1)
write.csv(low.expr.genenames, "~/Downloads/Rshiny_cell-markers/Rtables/hsc_ark_consistent_genes.csv",
          row.names=FALSE)

genelist <- c("BCMA", "CD38", "CD56", "CD138", "TACI", 
                 "SLAMF7", "CD19", "LY9", "ITGB7", "GPRC5D", 
                 "CD40", "CD44", "CD47", "BST2", "IGF1R", 
                 "ICAM1", "CD74", "CD81", "CD86", "CD200", "IGKC",
                 "FUT3", "CD70", "CD1D", low.expr.genenames)
# subset genes: get indices
idxs <- sapply(genelist, function(x) grep(paste0("^", x, "$"), rownames(hsc)))
hsc.mat <- hsc[idxs, ]
idxs <- sapply(genelist, function(x) grep(paste0("^", x, "$"), rownames(ark)))
ark.mat <- ark[idxs, ]

all.equal(row.names(hsc.mat), row.names(ark.mat))

#### now try z-score - use mean of "constant" genes
# hsc.zscore <- scale(hsc.mat, center=TRUE)
# (hsc.mat[3, 1] - mean(hsc.mat[, 1]))/sd(hsc.mat[, 1])
# head(hsc.zscore)
# # hist(hsc.zscore["BCMA", ], breaks=4)

# (hsc.mat[2, 1] - mean(hsc.mat[idxs, 1]))/sd(hsc.mat[idxs, 1])
# head((hsc.mat - mean(hsc.mat[idxs, ]))/sd(hsc.mat[idxs, ]))
idxs <- sapply(low.expr.genenames, function(x) grep(paste0("^", x, "$"), rownames(hsc.mat)))
hsc.zscore <- (hsc.mat - mean(hsc.mat[idxs, ]))/sd(hsc.mat[idxs, ])
# t.hsc <- t(hsc.mat)
# head((t.hsc - mean(t.hsc))/sd(t.hsc))

# ark.zscore <- scale(ark.mat, center=TRUE)
# ark.zscore[1:5, 1:5]
# # hist(ark.zscore["BCMA", ])
idxs <- sapply(low.expr.genenames, function(x) grep(paste0("^", x, "$"), rownames(ark.mat)))
ark.zscore <- (ark.mat - mean(ark.mat[idxs, ]))/sd(ark.mat[idxs, ])

### calc means
hsc.means <- apply(hsc.zscore, 1, mean)
ark.means <- apply(ark.zscore, 1, mean) 

### calc mm-normal
ark.hsc <- NULL
for(g in names(hsc.means)) {
  mm.norm <- mmrf.means[g] - hsc.means[g]
  mmrf.hsc <- c(mmrf.hsc, mm.norm)
  
  mm.norm <- ark.means[g] - hsc.means[g]
  ark.hsc <- c(ark.hsc, mm.norm)
}

# hist(ark.hsc)
summary(ark.hsc)
# barplot(ark.hsc)

head(ark.hsc)

### keep 24 genes
genelist <- c("BCMA", "CD38", "CD56", "CD138", "TACI", 
                 "SLAMF7", "CD19", "LY9", "ITGB7", "GPRC5D", 
                 "CD40", "CD44", "CD47", "BST2", "IGF1R", 
                 "ICAM1", "CD74", "CD81", "CD86", "CD200", "IGKC",
                 "FUT3", "CD70", "CD1D")
# subset genes: get indices
idxs <- sapply(genelist, function(x) grep(paste0("^", x, "$"), names(ark.hsc)))
ark.hsc.genes24 <- ark.hsc[idxs]
# barplot(ark.hsc.genes24)

### plot
mat <- ark.hsc.genes24
df <- data.frame(Genes=names(mat), Diff=mat)
df$Color <- ifelse(df$Diff > 0, "#CA0020", "#0571B0") # red for positive difference, blue for negative difference
df$Genes <- reorder(df$Genes, df$Diff) # order by difference
## plot
ggplot(df, aes(x=Genes, y=Diff)) +
  geom_bar(stat="identity", fill=df$Color) +
  coord_flip() +
  theme_pubr() +
  xlab("") +
  ylab("Cancer - Normal (z-score)")
```

```{r try (3)}
library(sva)

hsc <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/normals/gse17054_annot_log2.rds")
ark <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/mm_pts/arkansas_annot_log2.rds")

combined.mat <- cbind(hsc, ark)
combined.mat[1:5, 1:5]
batch <- factor(c(rep("hsc", ncol(hsc)), rep("mm", ncol(ark))))

modcombat <- model.matrix(~1, data=batch)
combat.edata <- ComBat(dat=combined.mat, batch=batch, mod=modcombat, par.prior=TRUE, prior.plots=FALSE)
```


Step 1
```{r load data}
hsc <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/normals/genes24_gse17054.rds")
hsc[1:5, 1:4] # 24 x 4
hist(hsc["BCMA", ], breaks=4)

mmrf <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/genes21/genes24_mmrf.rds")
mmrf[1:5, 1:5] # 24 x 598
hist(mmrf["BCMA", ])

ark <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/genes21/genes24_arkansas.rds")
ark[1:5, 1:5]
hist(ark["BCMA", ])
```

Step 2
```{r z-score}
hsc.zscore <- scale(hsc, center=TRUE)
head(hsc.zscore)
hist(hsc.zscore["BCMA", ], breaks=4)

ark.zscore <- scale(ark, center=TRUE)
ark.zscore[1:5, 1:5]
hist(ark.zscore["BCMA", ])

mmrf.zscore <- scale(mmrf)
# mmrf.zscore[1:5, 1:5]
# hist(mmrf.zscore["BCMA", ])
```

Step 3
```{r}
hsc.means <- apply(hsc.zscore, 1, mean)
ark.means <- apply(ark.zscore, 1, mean) 

# ### divide each vector by min value
# hsc.means <- hsc.means/min(hsc.means)
# ark.means <- ark.means/min(ark.means)
# 
# ### take negative inverse of any negative z-score means
# hsc.means[which(hsc.means < 0)] <- -1/hsc.means[which(hsc.means < 0)]
# ark.means[which(ark.means < 0)] <- -1/ark.means[which(ark.means < 0)]
# 
mmrf.means <- apply(mmrf.zscore, 1, mean)
```

Step 4 - calculate cancer minus normal
Want: High in cancer, low in normal -> value > 0
```{r}
mmrf.hsc <- NULL
ark.hsc <- NULL
for(g in names(hsc.means)) {
  mm.norm <- mmrf.means[g] - hsc.means[g]
  mmrf.hsc <- c(mmrf.hsc, mm.norm)
  
  mm.norm <- ark.means[g] - hsc.means[g]
  ark.hsc <- c(ark.hsc, mm.norm)
}

# hist(hsc.mmrf, breaks=8)
hist(ark.hsc)
summary(ark.hsc)

head(ark.hsc)
which(ark.hsc > 0.8683) # +1 std dev
which(ark.hsc < -0.7560) # -1 std dev

barplot(ark.hsc)
ark.hsc["IGKC"]
```

Step 5
```{r}
plot.title <- "MM-HSC difference, Arkansas"
filepath <- "~/Downloads/Rshiny_cell-markers/Rplots/figure3/HSC_MM_difference_Arkansas"

mat <- ark.hsc.genes24 ### ark.hsc or mmrf.hsc
df <- data.frame(Genes=names(mat), Diff=mat)
df$Color <- ifelse(df$Diff > 0, "#CA0020", "#0571B0") # red for positive difference, blue for negative difference

# genes.order <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/genes21/genes24_order.rds")
# df$Genes.relevel <- factor(df$Genes, levels=rev(genes.order)) # boxplot order
df$Genes <- reorder(df$Genes, df$Diff) # order by difference

tiff(filename=paste0(filepath, ".tiff"), res = 300, width = 6, height = 8, units = "in")
p <- ggbarplot(df, x="Genes", y="Diff", orientation="horiz",
               fill="Color", color="Color") 
ggpar(p, main=plot.title, xlab="", ylab="", legend="none")
dev.off()

saveRDS(df, file="~/Downloads/Rshiny_cell-markers/Rdata/plotting/df_hsc-vs-mm_ark.rds")

####### ggplot
ggplot(df, aes(x=Genes, y=Diff)) +
  geom_bar(stat="identity", fill=df$Color) +
  coord_flip() +
  theme_pubr() +
  xlab("") +
  ylab("Cancer - Normal (z-score)")


```