---
title: "BMCA low"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Identify patients with low BCMA expression (bottom third)
Note: don't make data into a dataframe before plotting histograms
```{r}
genes20.mmrf <- readRDS(file="~/Downloads/Rshiny_cell-markers/Rdata/genes21/genes24_mmrf.rds")
dim(genes20.mmrf) # 24 598
head(genes20.mmrf[, 1:6])
bcma <- genes20.mmrf["BCMA", ]
head(bcma)
hist(bcma, main="BCMA", xlab="log2(Gene expression)")
## get cutoff values to use to divide gene expression
# use +/- 1 standard deviation as cutoff
cutoff.low <- mean(bcma) - sd(bcma)
cutoff.hi <- mean(bcma) + sd(bcma)
bcma.low <- bcma[which(bcma < cutoff.low)]
length(bcma.low) # 79, 2 sd -> 20
bcma.med <- bcma[which(cutoff.low < bcma & bcma < cutoff.hi)]
length(bcma.med) # 443
bcma.hi <- bcma[which(cutoff.hi < bcma)]
length(bcma.hi) # 76, 2 sd -> 8

### Make plot
tiff("~/Downloads/Rshiny_cell-markers/Rplots/figure1/mmrf_BCMA_histogram.tiff", res = 300, width = 6, height = 4, units = "in")
hist(bcma, main="BCMA", xlab="log2(Gene expression)")
abline(v=cutoff.low, col="red")
abline(v=cutoff.hi, col="red")
dev.off()

### dataframe time
### make into dataframe
d <- data.frame(Genes=rownames(genes20.mmrf),
                genes20.mmrf)
rownames(d) <- NULL
str(d)
d[1:5, 1:5]

### just straight up do t-tests of BCMA-high vs -low?
bcma.factor <- factor(c(rep("low", length(bcma.low)), 
                        rep("medium", length(bcma.med)), 
                        rep("high", length(bcma.hi))),
                      levels=c("low", "medium", "high"))
names(bcma.factor) <- c(names(bcma.low), names(bcma.med), names(bcma.hi))
bcma.factor <- bcma.factor[order(names(bcma.factor))] # order patients
all.equal(names(d)[-c(1)], names(bcma.factor)) # TRUE
saveRDS(bcma.factor, file="~/Downloads/Rshiny_cell-markers/Rdata/factors/mmrf_bcma_levels.rds")

### t-test
t.test(d[d$Genes=="ICAM1", names(d) %in% names(bcma.factor[which(bcma.factor=="low")])],
       d[d$Genes=="ICAM1", names(d) %in% names(bcma.factor[which(bcma.factor=="high")])])


### aggregate bmca - turn into categorical variable
dd <- gather(d, key="Patient", value="Expression", -Genes)
head(dd)

bcma.level <- data.frame(Patient=c(names(bcma.low), 
                                   names(bcma.med), 
                                   names(bcma.hi)), 
                       BCMA.level=factor(c(rep("low", length(bcma.low)), 
                                    rep("medium", length(bcma.med)), 
                                    rep("high", length(bcma.hi))),
                                    levels=c("low", "medium", "high")
                                    )
                       )
head(bcma.level)
table(bcma.level$BCMA.level)

dd.bcma <- merge(dd, bcma.level, by="Patient")
head(dd.bcma)
saveRDS(dd.bcma, file="~/Downloads/Rshiny_cell-markers/Rdata/mmrf_dd_bcma_levels.rds")


### Save plot
tiff("~/Downloads/Rshiny_cell-markers/Rplots/figure1/BCMA_boxplot_genes24.tiff", res = 300, width = 10, height = 7, units = "in")
## specify how many groups need color
ncols <- 3
mycolors <- colorRampPalette(brewer.pal(9, "Blues"))(ncols)
## provide gene order from make_boxplots
genes.order <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/genes21/genes24_order.rds")
dd.bcma$Genes.relevel <- factor(dd$Genes, levels=genes.order)
## plot
ggplot(dd.bcma, aes(x=Genes.relevel, y=Expression, fill=BCMA.level)) +
  geom_boxplot(notch=TRUE, outlier.shape=NA, show.legend=TRUE) +
  scale_fill_manual(values=mycolors) +
  theme(axis.text.x = element_text(angle=90, size=10)) +
  theme(axis.title = element_text(size=12)) +
  ggtitle("Gene expression stratified by BCMA expression - MMRF") + 
  ylab("log2(Expression)") +
  xlab("Genes")
dev.off()
```

<!--
Calculate the statistics - ANOVA followed by Tukey
```{r}
### ANOVA
head(bcma.factor)
## make sure the following is true before running
all.equal(colnames(genes20.mmrf), names(bcma.factor)) # TRUE
markers <- rownames(genes20.mmrf)
for(gene in markers) {
  gene.expr <- genes20.mmrf[gene, ]
  
  # ANOVA
  model <- aov(gene.expr ~ bcma.factor)
  aov.pval <- summary(aov(model))[[1]][["Pr(>F)"]][[1]]
  
  # if significant, then Tukey
  if(aov.pval <= 0.05) {
    print(paste("yes", gene))
    
    tukey <- TukeyHSD(model)
    
    write.csv(tukey$bcma.factor, file=paste0("~/Downloads/Rshiny_cell-markers/Rtables/BCMA_mmrf_tukey/", gene, ".csv"))
  }
}

## t-test for sanity check on Tukey values
t.test(genes20.mmrf["BST2", which(bcma.factor=="low")], genes20.mmrf["BST2", which(bcma.factor=="low")])
```
-->

This isn't quite correct. It calculates difference between genes and difference between BCMA level, but can't look at difference between genes based on BCMA level. The interaction term is a little off.
<!--
```{r}
head(dd.bcma)
### aov
## Randomized Block Design (B is the blocking factor)?
## not for this data
# fit <- aov(Expression ~ Genes + BCMA.level, data=dd.bcma)
# summary(fit)

## Two Way Factorial Design?
## looking at interaction terms while controlling for genes and BCMA status
fit <- aov(Expression ~ Genes + BCMA.level + Genes*BCMA.level, data=dd.bcma)
summary(fit)

## Tukey post-hoc test
tukey.bcma <- TukeyHSD(fit)
saveRDS(tukey.bcma, file="~/Downloads/Rshiny_cell-markers/Rdata/mmrf_bcma_tukey.rds")
head(tukey.bcma$`Genes:BCMA.level`)
write.csv(tukey.bcma$`Genes:BCMA.level`, file="~/Downloads/Rshiny_cell-markers/Rtables/mmrf_bcma_tukey.csv")
## pull out information in Excel by hand
```
-->

BCMA stratification on Arkansas patients
```{r}
genes20.arkansas <- readRDS(file="~/Downloads/Rshiny_cell-markers/Rdata/genes21/genes24_arkansas.rds")
dim(genes20.arkansas) # 24 1035
head(genes20.arkansas[, 1:6])

### read in baseline/relapse annotation
arkansas.condition <- readRDS(file="~/Downloads/Rshiny_cell-markers/Rdata/factors/arkansas_baseline_relapse.rds")
head(arkansas.condition)

all.equal(colnames(genes20.arkansas), names(arkansas.condition)) # TRUE 
## line above must be true in order to run this:
bcma <- genes20.arkansas["BCMA", which(arkansas.condition=="baseline")]
bcma <- genes20.arkansas["BCMA", ] ## all Arkansas patients
length(bcma) # 255
hist(bcma, main="BCMA", xlab="log2(Gene expression)") # There's one patient with really low BCMA expression

## get cutoff values to use to divide gene expression
# use +/- 1 standard deviation as cutoff
cutoff.low <- mean(bcma) - sd(bcma)
cutoff.hi <- mean(bcma) + sd(bcma)
bcma.low <- bcma[which(bcma < cutoff.low)]
length(bcma.low) # 25
bcma.med <- bcma[which(cutoff.low < bcma & bcma < cutoff.hi)]
length(bcma.med) # 212
bcma.hi <- bcma[which(cutoff.hi < bcma)]
length(bcma.hi) # 18

### Make plot
tiff("~/Downloads/Rshiny_cell-markers/Rplots/figure1/Arkansas_all_BCMA_histogram.tiff", res = 300, width = 6, height = 4, units = "in")
hist(bcma, main="BCMA", xlab="log2(Gene expression)")
abline(v=cutoff.low, col="red")
abline(v=cutoff.hi, col="red")
dev.off()

### dataframe time
### make into dataframe
## subset patients who are relapsed
d <- data.frame(Genes=rownames(genes20.arkansas),
                genes20.arkansas[, which(arkansas.condition=="baseline")])
d <- data.frame(Genes=rownames(genes20.arkansas),
                genes20.arkansas) ## all Arkansas patients
rownames(d) <- NULL
d[1:5, 1:5]

### make factor levels
bcma.factor <- factor(c(rep("low", length(bcma.low)), 
                        rep("medium", length(bcma.med)), 
                        rep("high", length(bcma.hi))),
                      levels=c("low", "medium", "high"))
names(bcma.factor) <- c(names(bcma.low), names(bcma.med), names(bcma.hi))
bcma.factor <- bcma.factor[order(names(bcma.factor))] # order patients
all.equal(names(d)[-c(1)], names(bcma.factor)) # TRUE
saveRDS(bcma.factor, file="~/Downloads/Rshiny_cell-markers/Rdata/factors/arkansas_all_bcma_factors.rds")


### aggregate bmca - turn into categorical variable
dd <- gather(d, key="Patient", value="Expression", -Genes)
head(dd)

bcma.level <- data.frame(Patient=c(names(bcma.low), 
                                   names(bcma.med), 
                                   names(bcma.hi)), 
                       BCMA.level=factor(c(rep("low", length(bcma.low)), 
                                    rep("medium", length(bcma.med)), 
                                    rep("high", length(bcma.hi))),
                                    levels=c("low", "medium", "high")
                                    )
                       )
head(bcma.level)
table(bcma.level$BCMA.level)

dd.bcma <- merge(dd, bcma.level, by="Patient")
head(dd.bcma)
saveRDS(dd.bcma, file="~/Downloads/Rshiny_cell-markers/Rdata/Arkansas_all_dd_bcma_levels.rds")


### Save plot
tiff("~/Downloads/Rshiny_cell-markers/Rplots/figure1/Arkansas_all_BCMA_boxplot_genes24.tiff", res = 300, width = 10, height = 7, units = "in")
## specify how many groups need color
ncols <- 3
mycolors <- colorRampPalette(brewer.pal(5, "Blues"))(ncols)
## provide gene order from make_boxplots
genes.order <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/genes21/genes24_order.rds")
dd.bcma$Genes.relevel <- factor(dd$Genes, levels=genes.order)
## plot
ggplot(dd.bcma, aes(x=Genes.relevel, y=Expression, fill=BCMA.level)) +
  geom_boxplot(notch=TRUE, outlier.shape=NA, show.legend=TRUE) +
  scale_fill_manual(values=mycolors) +
  theme(axis.text.x = element_text(angle=90, size=10)) +
  theme(axis.title = element_text(size=12)) +
  ggtitle("Gene expression stratified by BCMA expression - Arkansas pts") + 
  ylab("log2(Expression)") +
  xlab("Genes")
dev.off()
```

### Boston
BCMA stratification on Boston patients
```{r boston}
genes20.boston<- readRDS(file="~/Downloads/Rshiny_cell-markers/Rdata/genes21/genes24_boston.rds")
dim(genes20.boston) # 24 264
head(genes20.boston[, 1:6])

bcma <- genes20.boston["BCMA", ]
length(bcma) # 264
hist(bcma, main="BCMA", xlab="log2(Gene expression)")

## get cutoff values to use to divide gene expression
# use +/- 1 standard deviation as cutoff
cutoff.low <- mean(bcma) - sd(bcma)
cutoff.hi <- mean(bcma) + sd(bcma)
bcma.low <- bcma[which(bcma < cutoff.low)]
length(bcma.low) # 34
bcma.med <- bcma[which(cutoff.low < bcma & bcma < cutoff.hi)]
length(bcma.med) # 192
bcma.hi <- bcma[which(cutoff.hi < bcma)]
length(bcma.hi) # 38

### Make plot
tiff("~/Downloads/Rshiny_cell-markers/Rplots/figure1/Boston_BCMA_histogram.tiff", res = 300, width = 6, height = 4, units = "in")
hist(bcma, main="BCMA", xlab="log2(Gene expression)", breaks=20)
abline(v=cutoff.low, col="red")
abline(v=cutoff.hi, col="red")
dev.off()

### dataframe time
### make into dataframe
d <- data.frame(Genes=rownames(genes20.boston),
                genes20.boston)
rownames(d) <- NULL
d[1:5, 1:5]

### just straight up do t-tests of BCMA-high vs -low?
bcma.factor <- factor(c(rep("low", length(bcma.low)), 
                        rep("medium", length(bcma.med)), 
                        rep("high", length(bcma.hi))),
                      levels=c("low", "medium", "high"))
names(bcma.factor) <- c(names(bcma.low), names(bcma.med), names(bcma.hi))
bcma.factor <- bcma.factor[order(names(bcma.factor))] # order patients
all.equal(names(d)[-c(1)], names(bcma.factor)) # TRUE
saveRDS(bcma.factor, file="~/Downloads/Rshiny_cell-markers/Rdata/factors/boston_bcma_levels.rds")

### t-test
t.test(d[d$Genes=="BCMA", names(d) %in% names(bcma.factor[which(bcma.factor=="low")])],
       d[d$Genes=="BCMA", names(d) %in% names(bcma.factor[which(bcma.factor=="high")])])


### aggregate bmca - turn into categorical variable
dd <- gather(d, key="Patient", value="Expression", -Genes)
head(dd)

bcma.level <- data.frame(Patient=c(names(bcma.low), 
                                   names(bcma.med), 
                                   names(bcma.hi)), 
                       BCMA.level=factor(c(rep("low", length(bcma.low)), 
                                    rep("medium", length(bcma.med)), 
                                    rep("high", length(bcma.hi))),
                                    levels=c("low", "medium", "high")
                                    )
                       )
head(bcma.level)
table(bcma.level$BCMA.level)

dd.bcma <- merge(dd, bcma.level, by="Patient")
head(dd.bcma)
saveRDS(dd.bcma, file="~/Downloads/Rshiny_cell-markers/Rdata/boston_dd_bcma_levels.rds")


### Save plot
tiff("~/Downloads/Rshiny_cell-markers/Rplots/figure1/Boston_BCMA_boxplot_genes24_theme-bw.tiff", res = 300, width = 10, height = 7, units = "in")
## specify how many groups need color
ncols <- 3
mycolors <- colorRampPalette(brewer.pal(9, "Blues"))(ncols)
## provide gene order from make_boxplots
genes.order <- readRDS("~/Downloads/Rshiny_cell-markers/Rdata/genes21/genes24_order.rds")
dd.bcma$Genes.relevel <- factor(dd$Genes, levels=genes.order)
## plot
ggplot(dd.bcma, aes(x=Genes.relevel, y=Expression, fill=BCMA.level)) +
  geom_boxplot(notch=TRUE, outlier.shape=NA, show.legend=TRUE) +
  scale_fill_manual(values=mycolors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, size=10)) +
  theme(axis.title = element_text(size=12)) +
  ggtitle("Gene expression stratified by BCMA expression - Boston") + 
  ylab("log2(Expression)") +
  xlab("Genes")
dev.off()

```


<!--
Calculate the statistics - ANOVA followed by Tukey
```{r}
### ANOVA
head(bcma.factor)
genes20.arkansas.relapse <- genes20.arkansas[, which(arkansas.condition$Condition=="relapse")]
dim(genes20.arkansas.relapse) # 24 255
## make sure the following is true before running
all.equal(colnames(genes20.arkansas.relapse), names(bcma.factor)) # TRUE
markers <- rownames(genes20.arkansas.relapse)
for(gene in markers) {
  gene.expr <- genes20.arkansas.relapse[gene, ]
  
  # ANOVA
  model <- aov(gene.expr ~ bcma.factor)
  aov.pval <- summary(aov(model))[[1]][["Pr(>F)"]][[1]]
  
  # if significant, then Tukey
  if(aov.pval <= 0.05) {
    print(paste("yes", gene))
    
    tukey <- TukeyHSD(model)
    
    write.csv(tukey$bcma.factor, file=paste0("~/Downloads/Rshiny_cell-markers/Rtables/BCMA_arkansas_tukey/", gene, ".csv"))
  }
}

## t-test for sanity check on Tukey values
t.test(genes20.arkansas.relapse["GPRC5D", which(bcma.factor=="low")], genes20.arkansas.relapse["GPRC5D", which(bcma.factor=="high")])
```
-->